---
title: "lab_05"
author: "Sean Mussenden"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and janitor.

**Task** load these two packages.

```{r}
# Turn off scientific notation
options(scipen=999)

library(tidyverse)
library(lubridate)
library(refinr)
```

## Load Data

You'll need to load three data sets for this:

* The West Virginia slice of the PPP loan data (lab_05.rds).
* A "lookup table" that allows you to translate NAICS (industry) numeric codes to industry titles (naics_codes.csv).
* A table of West Virginia population by county (American Community Survey, 2019 5-year averages) (wv_population_county.csv).

All three data sets are in the data folder.  Write code to load the three in the codeblock below.


```{r}
wv_ppp_loans <- read_rds("data/lab_05.rds")
naics_codes <- read.csv("data/naics_codes.csv")
wv_population_county <- read.csv("data/wv_population_county.csv")
```


**Task** Create a codeblock below this task, then read the data sets in in and assign them to appropriate variable names. There's a fourth data set you'll use in this lab, too, of selected loans in ZIP Code 25401. But there's no need to load it now.

## Answer questions

**Q1.** In the data folder, there is a csv called zip_25401_loan_sample.csv.  It contains a sample of loans from West Virginia ZIP Code 25401. 

As we read earlier this semester, [multiple loan applications coming from multiple businesses at the same residential street address](https://www.nytimes.com/2021/08/17/business/ppp-fraud-covid.html) might point to fraud. Or it could alert us to companies that used [multiple corporate entities to get more money than envisioned](https://www.propublica.org/article/different-names-same-address-how-big-businesses-got-government-loans-meant-for-small-businesses) by the original law that authorized the program.   

You are going to examine this data to see if we can find a lot of loans coming from the same address.  Here's the problem: the street address field is pretty messy.  The same address appears with minor variations --  "1003 Sushruta Dr" vs "1003 SUSHRUTA DR" -- that will prevent proper grouping. 

First, upload the data into Open Refine and standardize/clean the address field. If you've done it properly, you should have 65 discrete addresses. 

Then export the data from Open Refine, and move it to the proper folder. 

Next, load it in the codeblock below, assigning it to an appropriate variable name. 

Then answer these questions:
* What is the street address in this data that has the most loans?
* How many loans are there at that street address?
* What are the names of the businesses at that address?

**A1.** The address with the most loans is 126 E Burke St., which lists six loans. Businesses at the address are Drew Holdings, Brix27, Abraham Ashton, Hub Co-op, and Ronin Properties. (Bonus trivia: 126 E Burke St. is the site of the girlhood home of confederate spy Belle Boyd. )

```{r}
zip_25401_loans_address_cleaned <- read_csv("data/zip_25401_loan_sample_ORcleaned.csv")

zip_25401_loans_address_cleaned %>%
  group_by(address) %>%
  summarize(num_biz = n()) %>%
  arrange(desc(num_biz))

zip_25401_loans_address_cleaned %>%
  filter(address == "126 E Burke St")
```

**Q2.** Do some web research on the businesses that answered question 1.  

Google the street address.  Google the business names and search for their corporate records on [Open Corporates](https://opencorporates.com/). Be sure to find the website of the name of the company that appears twice in the list of businesses and develop an understanding of what it does. 

Based on your research, does it seem suspicious that this collection of businesses all got loans using the same address? Why or why not. Use specific facts identified in your research to support your position. 

**A2.** Business-by-business rundown of my findings: 

1. Drew Holdings: Its website contains almost nothing -- no rental properties, no documents. It's listed in OpenCorporates as a 4-year-old company (2017) with Abraham Ashton and Robert Johnson as agent and directors/officers. The website has an email address for Jaz Properties, which did not apply for a PPP loan but is listed in OpenCorporates as a real estate company at the Burke Street address, with Robert Johnson as agent. Jaz Properties' license had been revoked for failure to file an annual report. More on Drew Holdings below. 

2. Ronin Properties: OpenCorporates also lists Abraham Ashton and Robert Johnson as leaders of this company. It has no website. (The only Ronin Property Management website I could find was in Milwaukee.) The West Virginia tax assessors office shows it owns two properties in Berkeley County -- 315 Gussie Ave, and 300 N High st. More on WV tax assessors below.

3. Hub Co-op, which got two loans, appears to be a public business at this address and appears to operate as a WeWork-style shared work space operation. Per the West Virginia tax assessor's office (https://mapwv.gov/Assessment/Detail/?PID=02060014048200000000), Hub Co-op has owned the Burke St property since 2017, when it was purchased from Jaz Properties, who is listed as the 2016 owner. (Coincidence?) Abraham Ashton is listed as a director/officer of Hub Co-op. Robert Johnson had been an officer, but was removed in mid-2019. 

4. Abraham Ashton, the business: This is likely his law practice. OpenCorporates lists Ashton & Crawley-Woods legal services firm at the Burke address, but the license had been terminated (2019). OpenCorporates also lists The Ashton Law Firm with the Burke St address. LinkedIn surfaced a different Martinsburg address for The Ashton Law Firm at 208 S Queen St. Per the West Virginia State Bar, there is an Abraham Ashton with an active law license in West Virginia at the 208 S Queen St address. Google shows the WeWork building when you search for The Ashton Law Firm. The Ashton Law Firm at 208 S Queen St (a block away) is labeled as "closed" on Google Maps. To summarize, I think Abraham Ashton's law work happens at the Burke St address.

5. Brix27: An actual restaurant on 131 N Queen St. OpenCorporate has Robert Johnson and Abraham Ashton listed as officers. Drew Holdings is an officer. 

Per the West Virginia tax assessor, Drew Holdings owns the building at 131-135 Queen where the restaurant is located and two other Queen St properties (142 N and 144 N). In total, Drew Holdings owns 15 properties in West Virginia -- 10 in Berkeley County, 4 in Jefferson County, one in Morgan County.

[Digression. There are 63 businesses in OpenCorporates where an Abraham Ashton is listed as the officer of a West Virginia company. I didn't look at them all. Some appear unrelated to the Martinsburg Ashton. Interesting ones include Mountaineer Floor Cleaners (Ashton and Johnson again, company terminated), and Fuzzy Tail at the 126 Burke Street address (more real estate).]

Cross referencing the above with reality ... Drew Johnson is described in a Journal News article (https://www.journal-news.net/journal-news/future-of-the-old-post-office-federal-building-in-the-works-with-its-new-owner/article_ecdd48b3-1896-5045-be6c-8f77308b31cf.html) as sole owner of Drew Holdings LLC, and is described as purchasing and restoring Martinsburg buildings for business development. OpenCorporates does not list a Drew Johnson in Martinsburg, WV, at all, and his name is not on the OpenCorporates listing for Drew Holdings. But that's because he is Robert A. "Drew" Johnson where Drew is short for Andrew, thank you Facebook. (https://www.facebook.com/TheJournalWV/posts/robert-a-drew-johnson-owner-of-brix-27-said-the-restaurant-and-retail-store-will/1842177082473833/)

So, my working understanding is that Robert "Drew" Johnson and Abraham Ashton have some kind of lawyerly / real estate business relationship. Drew Holdings is the entity that purchases commercial real estate. Ronin Properties fulfills a similar purpose but for residential rentals. They have an actual restaurant that employs people, as well as the rental work space. They use the Hub Co-op as their business address and probably have their business office(s) in the building. Despite initial appearances, these all appear to be legitimate businesses. I'd want to double check my sleuthing with some phone calls. But, the loan amounts do not seem extreme for keeping restaurant employees on the payroll, as well as keeping some administrative people on board to handle real estate paperwork and maintenance. Not suspicious.



**Q3.** Start by using the West Virginia slice of the PPP loan data that you loaded at the start of the lab to create a subset of PPP loans in West Virginia's second largest county (which you can find in wv_population_county). And then use that table you created to answer the following questions:

* Which city in that county had the highest number of loans? 
* In that city, which industry title had more loans than any other industry title? 

Requirement: you MUST export a dataframe of PPP loans from R Studio at some point in the process (not necessarily at the beginning!), load it into Open Refine, clean the city column, export it from Open Refine, and reimport into R Studio. To export data, you will use the write_csv() function.

Guidance: there are a lot of steps you'll need to take to answer this question. You may or may not find it helpful to write out in English what you plan to do step-by-step before you start writing code.   

**A3.** Martinsburg had the highest number of loans. Full service restaurants had more loans than any other industry. 

```{r}
wv_population_county %>%
  arrange(desc(population_2019))

berkeley_wv_loans <- wv_ppp_loans %>%
  filter(project_county_name == "BERKELEY")

write.csv(berkeley_wv_loans,"data/berkeley_wv_loans.csv")

berkeley_wv_loans_citycleaned <- read_csv("data/berkeley_wv_loans_city_ORcleaned.csv")

berkeley_wv_loans_citycleaned %>%
  group_by(city) %>%
  summarize(num_berk_city_loans = n()) %>%
  arrange(desc(num_berk_city_loans))

colnames(naics_codes)
colnames(berkeley_wv_loans_citycleaned)

berkeley_clean_plus_naics <- berkeley_wv_loans_citycleaned %>% left_join(naics_codes, by=c("naics_code" = "ï..naics_code"))

berkeley_clean_plus_naics %>%
  filter(city == "Martinsburg") %>%
  group_by(title) %>%
  summarise(top_mrtnsbrg_biz = n()) %>%
  arrange(desc(top_mrtnsbrg_biz))
  
  


```

**Q4.** What are your two best hypotheses, which you'd need to confirm with reporting and further analysis, that explain why that industry is at the top of the list?

**A4.** Restaurants lost their customers immediately as full service dining became dangerous and health departments shut down dining. Even those restaurants that retooled for takeout needed time to change their business models. To stay open, and to give themselves time to adjust to pandemic food service, they needed help paying their employees. 

Also, compared to other industries, restaurants seem to operate on just-in-time cash flow basis. They need to purchase perishable food and pay cooks and servers as their business flows through. Unlike other industries, they don't seem as able to bank cash for a rainy day. 

```{r}
berkeley_restaurants <- berkeley_clean_plus_naics %>%
  filter(city == "Martinsburg") %>%
  filter(title == "Full-Service Restaurants")

berkeley_restaurants %>%
  filter(amount == max(amount))

```


**Q5.** Start with a table of loans to all businesses in the city and industry that answered question 3. Answer the following questions:
* What is the name of the business that got the highest approved loan amount? 
* How much was it for?
* When was it approved?
* How many jobs does the data say were retained?
* Is there a difference between the business' name in the PPP data and the name its customers know it by? If so, what is that name?
* How many locations does this business have? 
* Did one of its locations close during the pandemic, either before or after it got the loan?

Hint: you will not find the answers to the last three questions in the data.  You could call them directly to get that information, but I don't want you to do that for this assignment.  Instead, do some web research. I would start by Googling the company name from the data and looking at the page that comes up for the business from at http://apps.sos.wv.gov/. I would use information I found on that page and use info about the company from Google, the [Wayback machine](https://archive.org/web/) (which lets you look at older versions of a company's website), Yelp, and Facebook. 

**A5.**
Cottle Christi got the highest loan amount at $280,434. It was approved on Feb. 17, 2021. The data reports 94 jobs retained. The public knows the main restaurant as Kitzie's Restaurant & Lounge. Kitzie's appears to have two locations currently -- Spring Mills (Williamsport Pike) and Martinsburg (Rock Cliff). Riverbend Bar & Grill in Falling Waters also still appears to be in operation. 

The West Virginia business license shows six locations, two of them formally closed. Kitzie's Cafe II's license was terminated on 11/5/2018; Kitzie's of Inwood on 5/19/20 (just as the pandemic was gaining steam). (http://apps.sos.wv.gov/business/corporations/organization.aspx?org=338507) I can't find information about the closing of the original Kitzie's Cafe location. 

It's unclear when Kitzie's applied for the loan. If the implication is that Kitzie's got more money than she should have gotten, given that one location closed in May 2020, I'd want to know more about the size and staffing of the three restaurants still in operation before making that call. 
