---
title: "lab_03"
author: "sean mussenden"
date: "8/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here
library(tidyverse)

# Load janitor here
library(janitor)

```

## Load data
```{r}
# Write code to load the WV slice of the ppp loan data here and call it wv_ppp_loans.
# The data is called lab_03.rds, and is in the same folder as lab_03.Rmd (this data notebook)
wv_ppp_loans <- read_rds("lab_03.rds")

```

## Answer questions

Q1: Which five cities located in Kanawha County, West Virginia got the most loans (in number of loans)? How many did each get? To answer this question, you'll have to clean the very messy "city" field.

Note: To answer this question, please fully clean the city field for Kanawha County cities. Hint: once cleaned, the dataframe that counts the number of loans by city should have 38 distinct cities for Kanawha County.

A1: Charleston, South Charleston, Saint Albans, Dunbar and Nitro

```{r}

#glimpse(wv_ppp_loans)

city_cleaned_wv_ppp_loans <- wv_ppp_loans %>%
  mutate(city = str_to_title(city)) %>%
  mutate(city = case_when(
    city == "Balle" ~ "Belle",
    city == "Charelston" ~ "Charleston",
    city == "Crosslanes" ~ "Cross Lanes",
    city == "Elkveiw" ~ "Elkview",
    city == "S Charleston" ~ "South Charleston",
    city == "So Charleston" ~ "South Charleston",
    str_detect(city,"Alban") ~ "Saint Albans",
    str_detect(city,"alban") ~ "Saint Albans",
    TRUE ~ city
  )) 
  
city_cleaned_wv_ppp_loans %>%
  filter(project_county_name == "KANAWHA") %>%
  group_by(city) %>%
  summarise(
    count_loans=n()
  ) %>%
  arrange(desc(count_loans))



#mutate(city = str_to_title(city)) %>%
#  mutate(city = case_when(
#    city == "Anold" ~ "Arnold",
#    TRUE ~ city



```

Q2: What is the name of the business that got the largest loan (by dollar value) in the 5-digit ZIP code in West Virginia that had the highest dollar amount in total loans (by dollar value)? Write up a short paragraph that also answers these questions:

* Where in West Virginia is this ZIP code located?
* How much was the business' loan for?
* How much of that loan was forgiven?
* Which number is bigger, the loan amount or the forgiveness amount? Can you explain the discrepancy?

A2: Jackson Kelly PLLC

The West Virginia zip code with the highest total dollar amount in loans is 25301, located in Charleston. The biggest loan in the county, worth $5,068,472, went to the law offices of Jackson Kelly PLLC. The database reports that the entire loan value would be used for payroll. In June, $5,126,619 of the loan was forgiven, more than was approved. 

To understand the discrepancy between the amount forgiven and the total amount approved, it might be necessary to trace back the company's loan applications to see if there were multiple applications or an early application. The database also reports that the original loan amount was $2,880,600. If the law firm applied early in the loan program, the extra money forgiven may reflect accumulated interest on earlier loans. 

```{r}
# Hint
# First, create a dataframe that shows the top 5-digit ZIP code by total dollar amount

zip_wv_ppp_loans <- wv_ppp_loans %>%
  mutate(zip = str_sub(zip, start=1L, end=5L)) %>%
  group_by(zip) %>%
  summarise(zip_total_loans = sum(amount)) %>%
  arrange(desc(zip_total_loans)) %>%
  top_n(5)

# Second, use the top ZIP code to create a separate dataframe that shows the business with the biggest loan in that ZIP code.

topzip_wv_ppp_loans <- wv_ppp_loans %>%
  mutate(zip = str_sub(zip, start=1L, end=5L)) %>%
  filter(zip == 25301) %>%
  arrange(desc(amount))

wv_ppp_loans %>%
  filter(id == 94295751)
  

```

Q3: Five business entities owned by the spouse of a member of West Virginia's U.S. House delegation got PPP loans. Write up a paragraph that answers the following questions.   

  * What are the businesses' names?
  * What do the businesses do?
  * How much money did the five businesses receive in total?
  * What percentage of the total loan amount have been forgiven?
  * How did the U.S. House member vote on the legislation initially authorzing the PPP loan program.

Feel free to add any interesting "color" (little interesting details) you find in your research.  

Some hints to help you answer this question:

* West Virginia's three U.S. House members are listed [here](https://www.govtrack.us/congress/members/WV#representatives)
* To find businesses they or their families own, you can search for financial disclosure forms for U.S. House members [here](https://disclosures-clerk.house.gov/PublicDisclosure/FinancialDisclosure#Search).
  * To speed things up, here are direct links to the three U.S. House members most recent financial disclosures, which I found by searching the site linked above.
  * [Rep. 1](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10042042.pdf)
  * [Rep. 2](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10040330.pdf)
  * [Rep. 3](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10042734.pdf)
* When you examine the disclosure forms, there may be lots of stock sales and retirement accounts mixed in.  Ignore those. In the table in the disclosure forms, look for "Partnership" or "C-Corp" in the column named "Income Type"
* This is a link to details on passage of the [CARES ACT](https://www.congress.gov/bill/116th-congress/house-bill/748). You can find a link to how the House voted by clicking on a link from [this page](https://www.congress.gov/bill/116th-congress/house-bill/748/actions).

A3: Matt Miller, husband of Rep. Carol Miller, R-W.Va., got PPP loans for the family's chain of car dealerships (Dutch Miller of Charleston, Dutch Miller of Charlotte, Dutch Miller Subaru, Dutch Miller Chevrolet, and DM Motor). In total, the Miller family businesses got $3,105,300, all of it forgiven, with interest. 

Rep. Miller voted for the CARES Act that authorized PPP loans, as well as the two pieces of legislation that expanded the PPP funding. She later voted against the American Rescue Plan, which contained assistance to individuals in the form of child tax credits, mortgage and rental assistance, and extended unemployment benefits. She also voted against a $3.5 billion spending package, complaining on Twitter that it was "chocked full of socialist policies." That did not stop her for submitting earmark requests for local projects worth $2.7 million.

https://twitter.com/RepCarolMiller/status/1430290236903464969

```{r}

# Hint
# First, create a dataframe that only has loans to those five companies

miller_wv_ppp_loans <- wv_ppp_loans %>%
  filter(str_detect(name,"DM MOTOR")| str_detect(name,"DUTCH MILLER") | str_detect(name,"MFM")) 

# Second, use that newly created dataframe to calculate total loan amount and percentage forgiven for those companies

miller_wv_ppp_loans %>%
  summarise(
    miller_total = sum(amount),
    miller_total_forgive = sum(forgiveness_amount)
  ) %>%
  mutate(miller_pct_forgive = miller_total_forgive / miller_total)
  


```
