---
title: "lab_01"
author: "sean mussenden"
date: "8/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**). 
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**). 
* write out the answer in the form of a complete sentence in the space given (look for **A**). 
When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS. 

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse.To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}


# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data 

We're going to load the slice of PPP loans for West Virginia.

**Task** Run the codeblock below to load the data. 

```{r}
wv_ppp_loans <- read_rds("lab_01.rds")
```

## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data. 
**Q1** How many rows and columns does this dataframe have?
**A1** 33,703 rows, 63 columns

```{r}
glimpse(wv_ppp_loans)
```

## Answer questions

**Q2** Which West Virginia county got the most loans? How many?
**A2** Kanawha County, with 4,332 loans, got the most PPP loans in West Virginia.

```{r}
wv_ppp_loans %>%
  group_by(project_county_name) %>%
    summarise(
      count_wvloans = n()
      ) %>%
    arrange(desc(count_wvloans))
```
**Q3** Which West Virginia county got the fewest loans? How many?
**A3** Wirt County, with 36 loans, got the fewest.

```{r}
wv_ppp_loans %>%
  group_by(project_county_name) %>%
    summarise(
      count_wvloans = n()
      ) %>%
    arrange(count_wvloans)
```
**Q4** What's your best educated guess to explain why the county identified in question 2 had a lot of loans and why the county identified in question 3 had only a few loans? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.  
**A4** According to the U.S. Census Bureau, Kanawha County has the largest population and Wirt County has the smallest. (https://www.census.gov/programs-surveys/popest/technical-documentation/research/evaluation-estimates/2020-evaluation-estimates/2010s-counties-total.html)


**Q5** What West Virginia business got the state's largest PPP loan? How much was it for? What does the business do? How many jobs did the loan retain, according to the data?
**A5** The Greenbrier Hotel Corp. got West Virginia's largest loan, worth $8,912,202. The resort, a combined hotel, golf course, casino, conference center and former Cold War bunker, said it retained 500 jobs.

```{r}
wv_ppp_loans %>%
  arrange(desc(amount))

```

**Q6** What West Virginia politician (with his family) owns the company identified in question 2? What's your source for that information? Hint: this will require web research, you won't be able to answer this purely from the data set.  You do not need to write any code for this question, so no codeblock is provided.  
**A6** West Virginia Gov. Jim Justice and his family own the Greenbrier. (https://governor.wv.gov/about/Pages/default.aspx, https://www.bloomberg.com/profile/company/3288406Z:US)

**Q7**. Produce a summary table that has the following information for West Virginia as a whole (the whole state, not by county):
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings. 

**A7** West Virginian businesses got 33,703 loans, skewed heavily toward smaller-dollar loans. The median loan was $20,832, significantly smaller than the average loan of $77,338. The average loan's much smaller than the state's largest, nearly $9 million. West Virginia's economy appears to run on small businesses.
```{r}
wv_summary = wv_ppp_loans %>%
  summarise(
    count_loans = n(),
    total_loans = sum(amount),
    average_loan = mean(amount),
    median_loan = median(amount),
    smallest_loan = min(amount),
    largest_loan = max(amount)
  )
```

**Q8** Produce a summary table that has the following information for each West Virginia county:
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_county_summary` so that you can examine it with the data viewer. 

What newsworthy nuggets or patterns jump out at you that you could imagine dropping into a news story or doing follow up reporting on? Write up a few sentence that summarizes one county. 

**A8** Seven counties got fewer than 100 loans. Are those counties that small, or do businesses not know how or know to apply for the program? Do these smallest loans, as little as $165, make a difference in retaining jobs, the purpose of the PPP program? Calhoun, Webster and Dodridge Counties had fewer than 100 loans total but rank as the top 3 average loan size by county -- what's up with that? It would be interesting to map the average loan amount against population and find the outliers. 

Greenbrier County got the single largest loan, but the county's a smaller player in West Virginia's use of PPP loans. It's 908 total loans falls below average for West Virginia's counties, and even with its single $9 million loan skewing the stats, its average loan amount sits below the state's average.  
```{r}
wv_county_summary = wv_ppp_loans %>%
  group_by(project_county_name) %>%
  summarise(
    count_loans = n(),
    loans_total = sum(amount),
    average_loan = mean(amount),
    median_loan = median(amount),
    smallest_loan = min(amount),
    largest_loan = max(amount)
  ) %>%
  arrange(project_county_name)

```
